Rhyme: A Query Language for Nested Data Structures

### 2024-12-15 paper.tex
finish paper

@@

>>>>

In this section, we will compare our generated code for sparse tensor algebra with the state of the art sparse tensor compiler TACO. We will compare the generated code and execute them to compare the performance.\par
We use a simple tensor computation to compare the generated code - dot product of 4 sparse vectors. The mathematical definition of the computation is given below:
$$A = \sum_{i} B_i \cdot C_i \cdot D_i \cdot E_i$$
The query of this tensor computation for \rhyme and TACO are listed below:\par
In this section, we will compare the code generated by \rhyme for sparse tensor algebra with the state-of-the-art sparse tensor compiler, TACO. We will analyze the generated code and execute it to evaluate performance.\par
We use a simple tensor computation to compare the generated code: the dot product of four sparse vectors. The mathematical definition of this computation is given below:
$$A = \sum_{i} B_i \cdot C_i \cdot D_i \cdot E_i$$
The queries for this tensor computation in \rhyme and TACO are listed below:\par
\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
// Rhyme query
rh`sum(B.*i * C.*i * D.*i * E.*i)`
\end{lstlisting}
\end{minipage}%
\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
// TACO query
A = B(i) * C(i) * D(i) * E(i)
\end{lstlisting}

\vspace*{2mm}
\noindent
We will also provide additional type schema for the \rhyme query, indicating that the 4 vectors are sparse. The query will go through the type inference/checking, IR-generation, loop scheduling and code generation phase as we mentioned in Section \ref{workflow}. The final generated code for \rhyme and TACO are shown below (we only show the computation code and excluded the data loading and driver code).\par
\vspace*{2mm}
\noindent
\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
// Rhyme generated code
int tmp0 = 0;
for (auto xi_mit = CSVector<int, int>::multi_iterator({&B,&C,&D,&E});!xi_mit.finish(); ++xi_mit) {
  tmp0 += (int)(((*xi_mit).second[0] * (*xi_mit).second[1] * (*xi_mit).second[2] * (*xi_mit).second[3]));
}
\end{lstlisting}

\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
// TACO generated code
while (((i204B < pB1_end && i204C < pC1_end) && i204D < pD1_end) && i204E < pE1_end) {
  int32_t i204B0 = B1_crd[i204B];
  int32_t i204C0 = C1_crd[i204C];
  int32_t i204D0 = D1_crd[i204D];
  int32_t i204E0 = E1_crd[i204E];
  int32_t i204 = TACO_MIN(i204B0,TACO_MIN(i204C0,TACO_MIN(i204D0,i204E0)));
  if (((i204B0 == i204 && i204C0 == i204) && i204D0 == i204) && i204E0 == i204) {
    A0_val += ((B_vals[i204B] * C_vals[i204C]) * D_vals[i204D]) * E_vals[i204E];
  }
  i204B += (int32_t)(i204B0 == i204);
  i204C += (int32_t)(i204C0 == i204);
  i204D += (int32_t)(i204D0 == i204);
  i204E += (int32_t)(i204E0 == i204);
}
\end{lstlisting}
As we can observe, \rhyme 's generated code is more simple and can be directly mapped to the original query. \rhyme heavily utilize C++'s template feature and operator overloading to make the generated code more declarative and easy to understand. We also preserve the iteration logic in the original query by using the custom multi\_iterator. All the complex logic for co-advancing pointers in hidden in the header files.\par
As we can observe, \rhyme's generated code is simpler and can be directly mapped to the original query. \rhyme heavily utilizes C++'s template features and operator overloading to make the generated code more aligned with the original query logic and easier to understand. The iteration logic from the original query is preserved by using the custom \emph{multi\_iterator}. All the complex logic for co-advancing pointers is encapsulated within the header files.\par
      \includegraphics[scale=0.5]{figures/evaluation.pdf}
      \caption{\rhyme vs TACO}
      \label{evaluationpic}
We also compared the execution time of \rhyme and TACO, we run the generated code on 4 randomly generated sparse vectors with 10 percent sparsity. The result is shown in Figure \ref{evaluationpic}. As we can observe, \rhyme 's generated code can achieve roughly $80\%$ performance of TACO in the example query. The performance gap could be caused by the fact that TACO generates native C code for the computation kernel while our header file is purely written in C++ and heavily utilize STL containers. Our backend also adds several abstractions over both the sparse tensor data structure and the co-iteration method which could add additional overhead. But it should be noted that \rhyme is still an ongoing project and we could develop a pure C backend and generate more specialized C code with not much development effort. Also, compared with TACO, \rhyme manage to integrate relational algebra with sparse tensor algebra in a unified framework and perform several loop optimization likes loop hoisting and fusion.
We also compared the execution time of \rhyme and TACO by running the generated code on four randomly generated sparse vectors with 10$\%$ sparsity. The results are shown in Figure \ref{evaluationpic}. As observed, \rhyme's generated code achieves approximately $80\%$ of TACO's performance for the example query.\par
The performance gap may stem from the fact that TACO generates native C code for the computation kernel, while \rhyme's backend relies on C++ and extensively uses STL containers. Additionally, \rhyme introduces several abstractions over both the sparse tensor data structure and the co-iteration method, which could contribute to additional overhead.\par
However, it is important to note that \rhyme is still an ongoing project. With minimal development effort, we could implement a pure C backend to generate more specialized C code. Furthermore, compared to TACO, \rhyme integrates relational algebra and sparse tensor algebra into a unified framework while supporting advanced loop optimizations such as loop hoisting and fusion.

@@

In this paper, we present \rhyme - a declarative multi-paradigm query language that support both relational algebra and dense/sparse tensor algebra in a unified framework. We give a detailed explanation of how \rhyme supports specialized code generation for sparse tensor algebra and its type system design. \rhyme is still under development and we believe with more development effort, \rhyme can support multi-paradigm workload efficiently.
In this paper, we present \rhyme—a declarative multi-paradigm query language that supports both relational algebra and dense/sparse tensor algebra within a unified framework. We provide a detailed explanation of how \rhyme enables specialized code generation for sparse tensor algebra and the design of its type system. While \rhyme is still under active development, we believe that, with further effort, it can efficiently support multi-paradigm workloads.
